##
## PUBPOL543-Plot1: univariate - Frequency Table of Race in the United States
##  Menghan Zhai
##

upload ACS data
```{r}
library(foreign)
link = "https://github.com/G7-PUBPOL-543/ACS/raw/main/ManipulatedACS_V2.dta"
ACSData = read.dta(link)
```

check variable: race
```{r}
# Check names of all the variables in ACS Data
names(ACSData)
```
```{r}
#see values in the variable race:
head(ACSData$race,20)
```

get frequency table of race: RaceFreq
```{r}
# absolute values
RaceCount=table(ACSData$race,
                exclude = 'nothing') 
RaceCount
```
```{r}
# relative values
RaceProptoplot=prop.table(RaceCount)*100
RaceProptoplot
```
```{r}
# Turn the frequency table into a data frame:
RaceFreq=as.data.frame(RaceProptoplot)
```
```{r}
# renaming data frame
names(RaceFreq)=c("race","pct")

RaceFreq
```

Time to plot!
```{r}
library(ggplot2)
```

Order RaceFreq
```{r}
RaceFreq = RaceFreq[order(RaceFreq$pct),]

RaceFreq
```
```{r}
# save the order
RaceOrd = RaceFreq[order(RaceFreq$pct),'race']
```

base
```{r}
base = ggplot(data = RaceFreq, 
             aes(x = race,
                 y = pct)) 
base = base + theme_light() # add background: light theme
## add order
base= base + scale_x_discrete(limits=RaceOrd) 

base
```

plot1: + geom bar
```{r}
plot1 = base + geom_bar(stat='identity')

plot1
```

plot2: + title
```{r}
# name the title
TitleRace = list(Title = "Frequency Table of Race in the United States",
                 SubTi = NULL,
                 XTi = "Race",
                 YTi = "Percentage",
                 Sou = "Source: 2019 American Community Survey (ACS) Data")
```
```{r}
# add title
plot2 = plot1 + labs(title = TitleRace$Title,
                    subtitle = TitleRace$SubTi,
                    x = TitleRace$XTi,
                    y = TitleRace$YTi,
                    caption = TitleRace$Sou)
plot2
```

plot3: customize Y axis
```{r}
library(scales)
plot3 = plot2 + scale_y_continuous(breaks = c(5, 10, 20, 40, 60, 80),
                                   limits = c(0, 80),
                                   labels = unit_format(suffix = '%')) 
plot3
```

plot4: Title/subtitle positions:
```{r}
plot4 = plot3 + theme(plot.caption = element_text(hjust = 1),
                      plot.title = element_text(hjust = 0.5))
plot4 = plot4 + coord_flip() # flip the plot to fully show the types of race

plot4
```

plot5: Annotating the bars:
```{r}
paste0(round(RaceFreq$pct,2), '%')
LABELS=paste0(round(RaceFreq$pct,2), '%')
```
```{r}
plot5 = plot4 + geom_text(vjust = 0,
                          hjust = -0.1,
                          size = 3,
                          aes(y = pct ,
                              label = LABELS))
plot5
```

# Final Univariate Plot
```{r}
plot5
```



##
## PUBPOL543-Plot2: bivariate - Contingency Tables of Race and Health Insurance Coverage in the United States
## Menghan Zhai
##

I'll explore Race and Health Insurance Coverage in the United States in this plot.
As these two variable are both categorical, I'll use contigency tables to organize the relationship.
```{r}
head(ACSData,5)
```
```{r}
# check variable hcovany: with health insurance coverage or not
head(ACSData$hcovany,20)
```

```{r}
# contingency table of counts
(RaceHIC = table(ACSData$race,ACSData$hcovany)) # race and health insurance coverage
```
```{r}
# compute marginal percent (per column) from contingency table
library(magrittr)
(RaceHIC_mgCol = prop.table(RaceHIC,
                            margin = 2)%>%round(.,3))
```
```{r}
#make a data frame from contingency table
RaceHIC_DF = as.data.frame(RaceHIC)
names(RaceHIC_DF) = c("race","HICstatus","counts")

RaceHIC_DF
```
```{r}
#adding marginal percents:
RaceHIC_DF$pctCol = as.data.frame(RaceHIC_mgCol)[,3]

RaceHIC_DF
```

# a grouped chart
```{r}
library(ggplot2)
base1 = ggplot(data = RaceHIC_DF, 
             aes(x = HICstatus, y = counts,
                 fill= race ))

barGC = base1 +  geom_bar(stat = "identity",
                          position = 'dodge') 
barGC
```
```{r}
# add labels
barGC + geom_text(position = position_dodge(width=0.9),
                  angle = 90,
                  hjust = -0.5,
                  aes(label = counts)) 
```
```{r}
# change color
barGC + scale_fill_brewer(palette="Paired")
```

reorder by race
```{r}
RaceHIC_DF$race = factor(RaceHIC_DF$race,
                         levels = c("white","black/african american/negro","other asian or pacific islander", "other race, nec","two major races","chinese","american indian or alaska native","three or more major races","japanese"))
```

```{r}
library(ggplot2)
base1 = ggplot(data = RaceHIC_DF, 
               aes(x = HICstatus,
                   y = counts,
                   fill= race ))

barGC = base1 +  geom_bar(stat = "identity",
                          position = 'dodge') 
#labels
barGC = barGC + geom_text(position = position_dodge(width=0.9),
                          angle = 90,
                          hjust = -0.5,
                          aes(label = counts)) 
# palette with ordering
barGC = barGC + scale_fill_brewer(name = "PRECINCT",
                          palette = "BuPu",
                          direction = -1)
barGC
```

go from dodge to stack:
```{r}
# Stacked bar plot
conditionColor = ifelse(RaceHIC_DF$race%in%c("white",'black/african american/negro'),'grey80','grey50')
barStacked = base1 + geom_bar(stat = "identity",
                              position = 'stack')#default
barStacked = barStacked + geom_text(size = 5,
                                    fontface='bold',
                                    position = position_stack(vjust = 0.5),
                                    color=conditionColor,
                                    aes(label=counts))# its own AES!
barStacked + scale_fill_brewer(palette="GnBu",
                               direction = -1)
```

stacked percent
```{r}
library(scales) 

base2 = ggplot(data = RaceHIC_DF, 
               aes(fill = race,y = counts,x = HICstatus)) 

barStackPct = base1 + geom_bar(stat = "identity",
                               position = "fill")

barStackPct1 = barStackPct + geom_text(size = 5,
                                       position = position_fill(vjust = 0.5),
                                       aes(label = percent(pctCol,accuracy = 0.1)))

barStackPct1
```
```{r}
# change the ratio of x, y axis
barStackPct2 = barStackPct + geom_text(size = 1,
                                       position = position_fill(vjust = 0.5),
                                       aes(label = percent(pctCol,accuracy = 0.1)))

barStackPct2 = barStackPct2 + coord_fixed(ratio=10/1)+scale_y_continuous(breaks = seq(0,1,0.1))
barStackPct2 
```

Couldn't show all. It seems this is the "bad idea" introduced in class.
I would have to reproduce the table.

# try heatplot
```{r}
base  = ggplot(RaceHIC_DF, aes(x = HICstatus, 
                               y = reorder(race, pctCol), 
                               fill = pctCol*100)) 
heat = base +  geom_tile()

# color intensity
heat = heat + scale_fill_gradient(low = "white",
                                  high = "black")
heat = heat + theme_classic()

heat
```
```{r}
# improve heat plot

heat = heat + labs(y="Race", "Health Insurance Coverage Status")
heat = heat + theme(axis.text.x = element_text(angle = 0, 
                                               vjust = 0.6), 
                      legend.title = element_blank(), #no leg. title 
                      legend.position="top", 
                      legend.direction="horizontal",
                      legend.key.width=unit(1, "cm"),
                      legend.key.height=unit(1, "cm")) 

heat
```
```{r}
TitleRaceHIC = list(Title = "Heat Plot-Race and Health Insurance Coverage in the U.S.",
                    SubTi = NULL,
                    XTi = "Health Insurance Coverage Status",
                    YTi = "Race",
                    Sou = "Source: IPUMS USA")
```
```{r}
# add title
heat2 = heat + labs(title = TitleRaceHIC$Title,
                    subtitle = TitleRaceHIC$SubTi,
                    x = TitleRaceHIC$XTi,
                    y = TitleRaceHIC$YTi,
                    caption = TitleRaceHIC$Sou)
heat2
```
```{r}
# heat3: Title positions:
heat3 = heat2 + theme(plot.caption = element_text(hjust=1),
                      plot.title = element_text(hjust=0.5))

heat3
```
Still unobvious visually. Should I exclude "white"? Or is there some other way to retrieve this plot??

# try contingency table
```{r}
# plot a representation of contingency table:

library(ggplot2)                           
base3 = ggplot(RaceHIC_DF, aes(x = HICstatus,y = counts)) 
tablePlot = base3 + geom_point(aes(size = pctCol*100)) 
tablePlot = tablePlot + geom_text(aes(label = percent(pctCol)),
                                  nudge_x = 0.15,
                                  size = 3)
tablePlot
```
```{r}
# improve the previous plot
tablePlot = tablePlot + theme_minimal()
tablePlot = tablePlot + theme(legend.position = "none")

tablePlot
```
```{r}
# barplot with less information
base4 = ggplot(RaceHIC_DF, aes(x = race, y = counts)) 

# bars
bars = base4 + geom_bar(stat = "identity") + theme_minimal()

# bar per health insurance status with 'facet'
barsFa = bars + facet_grid(~ HICstatus) 

barsFa
```
```{r}
# improve previous plot
# change the minimal theme
barsFa = barsFa + theme(axis.text.x = element_text(angle = 90,
                                                   hjust = 1,
                                                   size = 7))
barsFa
```
```{r}
# original relationship Input-Output table
base5  = ggplot(RaceHIC_DF, aes(x = race,  y = pctCol ) ) 
barsIO = base5 + geom_bar( stat = "identity" )
barsIO = barsIO + facet_grid( ~ HICstatus) 
barsIO = barsIO + coord_flip()

barsIO
```

```{r}
# introduce "reorder""

# race ordered by pctcol
base5b  = ggplot(RaceHIC_DF, 
                 aes(x = reorder(race, pctCol), #here
                     y = pctCol) ) 

barsIOb = base5b + geom_bar(stat = "identity")
barsIOb = barsIOb + facet_grid( ~ HICstatus) 
barsIOb = barsIOb + coord_flip() 

barsIOb = barsIOb + theme(axis.text.y = element_text(size = 7,angle = 45)) 

barsIOb
```
Nice try. Next step is to:
add titles
```{r}
# name the title
TitleRaHI = list(Title = " Health Insurance Coverage Status by Race in the United States",
                 SubTi = NULL,
                 XTi = "Race",
                 YTi = "Percentage, Health Insurance Coverage Status",
                 Sou = "Source: 2019 American Community Survey (ACS) Data")
```
```{r}
# add title
barsIObTi = barsIOb + labs(title = TitleRaHI$Title,
                    subtitle = TitleRaHI$SubTi,
                    x = TitleRaHI$XTi,
                    y = TitleRaHI$YTi,
                    caption = TitleRaHI$Sou)
barsIObTi
```

title positions
```{r}
barsIObTi1 = barsIObTi + theme(plot.caption = element_text(hjust = 1),
                               plot.title = element_text(hjust = 0.5))
```

# Final Bivariable Plot
```{r}
barsIObTi1
```






##
## PUBPOL543-Map: Visualization of Poverty Rate by State in the United States
## Menghan Zhai
##

Process ACSData
```{r}
head(ACSData$poverty,40)
```

# generate a new povertyst variable to show if one is in poverty. According to IPUM.USA, poverty < 100 represents "in poverty" status, and poverty>=100 is regarded as "not in poverty".
```{r}
ACSData$povertyst 
ACSData$povertyst <- (ACSData$poverty < 100) # if poverty<100, then povertyst=TRUE, povertyst_num=1
```
```{r}
ACSData$povertyst_num <- as.numeric(ACSData$povertyst) # generate a numeric variable to calculate poverty rate by state
head(ACSData$povertyst_num,20)
```
```{r}
head(ACSData,5) #check
```
```{r}
library(dplyr)
ACSData1 <- data.frame(ACSData$statefip , ACSData$povertyst_num) # Create a new data frame
head(ACSData1,5)
```

# calculate the percent of "in poverty" of each state
generate "count" variable to count for the total number of person of each state
```{r}
ACSData1$count 
ACSData1$count <- 1
head(ACSData1,5)
```
calculate
```{r}
library(dplyr)
ACSData2 <- group_by(ACSData1, ACSData.statefip) %>%
# group_by(ACSData1, ACSData.STATEFP) %>% 
  mutate(percent = sum(ACSData.povertyst_num)/sum(count))
head(ACSData2,5)
```

save the "percent" to new dataframe
```{r}
ACSData3 <- ACSData2 %>%
  distinct(ACSData.statefip, percent, .keep_all = TRUE)

head(ACSData3,5)
```

# generate a new "NAME" variable  in ACSData3 which equals the original statefip in ACS data to merge the data and the map.
```{r}
ACSData3$NAME = ACSData3$ACSData.statefip

head(ACSData3,5)
```

# capitalize NAME in ACSData3 so that it is the same as NAME in map
```{r}
ACSData4 = as.data.frame(ACSData3) # change it to a data frame
head(ACSData4,5)
```
```{r}
ACSData4$NAME_char <- as.character(ACSData4$NAME) # Apply as.character function
head(ACSData4$NAME_char,5)
```
```{r}
library(Hmisc)
ACSData4$NAME_char = capitalize(ACSData4$NAME_char)
head(ACSData4,5)
```
```{r}
ACSData4$NAME <- as.factor(ACSData4$NAME_char) 
head(ACSData4,5)
```

# new column "poverty_rate" with the right name to show in map
```{r}
ACSData4$poverty_rate = ACSData4$percent 
head(ACSData4,5)
```

# open the map
```{r}
linkMap="https://github.com/G7-PUBPOL-543/maps/raw/main/states.geojson"
library(sf)
```
```{r}
mapUS=read_sf(linkMap)
```
```{r}
# see data in map
head(mapUS)
```

```{r}
str(ACSData4)
```

# Merge ACS data and U.S. map by state name
```{r}
mapUSVars=merge(mapUS,
                ACSData4, 
                by='NAME') 

mapUSVars
```

# 1.Base layer: Map of U.S.
```{r}
library(ggplot2)
# plot original map
base = ggplot(data = mapUS) + geom_sf(fill = 'grey90',
                                    color = NA) + theme_classic()

base #base map of the U.S.
```

# 2.Data layer: Poverty Rate Map by State in the United States
```{r}
povMap= base + geom_sf(data = mapUSVars,
                       aes(fill = poverty_rate),
                       color = NA)
povMap
```

change color
```{r}
povMap1 = povMap + scale_fill_gradient(low = 'blue',
                                       high= 'yellow')
```

add titles
```{r}
# name the title
TitlePovMap = list(Title = " Poverty Rate Map by State in the United States",
                   SubTi = NULL,
                   XTi = NULL,
                   YTi = NULL,
                   Sou = "Source: 2019 American Community Survey (ACS) Data")
```
```{r}
# add title
povMap2 = povMap1 + labs(title = TitlePovMap$Title,
                    subtitle = TitlePovMap$SubTi,
                    x = TitlePovMap$XTi,
                    y = TitlePovMap$YTi,
                    caption = TitlePovMap$Sou)
povMap2
```

title positions
```{r}
povMap3 = povMap2 + theme(plot.caption = element_text(hjust = 1),
                          plot.title = element_text(hjust = 0.5))

```
# Final Map
```{r}
povMap3
```


# That's all for my final project. Thank you!!